Probabilistic Programming around Gaussian Processes

### 2019-10-22 abstract/paper.tex
with bibrefs

Gaussian processes~\cite{} possess properties that make
them the approach of choice in many applications:
\begin{itemize}
    \item A Gaussian process works with as little or as much
    	data as available.
    \item Non-uniformly sampled observations, missing
    	observations, and observation noise are handled
    	organically.
    \item Libraries for inference with Gaussian processes
    	are readily available, both standalone and within
    	machine learning frameworks, such that Gaussian
    	processes can be included into data processing
    	pipelines and combined with other machine learning
    	algorithms.
\end{itemize}

Gaussian processes are successfully deployed in many settings,
in particular when 
\begin{itemize}
    \item a stationary kernel is sufficient;
    \item kernel hyperparameters can be fixed \textit{a priori} or
    	reliably estimated from observations;
    \item observation noise is constant.
\end{itemize}
Most basic cases are indeed covered by the above assumptions,
and library implementations of Gaussian processes prioritize
scalability with number of observations, rich libraries
of combinable kernel functions, and integration of Gaussian
processes with machine learning pipelines.

There are cases though where noise varies between
observations, structured informative priors must be imposed on
kernel hyperparameters, observation inputs are uncertain, or
observations are best explained by presence of change
points~\cite{}.

While many of the above cases are addressed in research
literature, their diversity is beyond feasible support by a
library, and the practitioner has to implement advanced Gaussian
process algorithms on case-by-case basis. One, laborious, option
is to implement the algorithms in a general-purpose programming
language, along with probabilistic inference for uncertain
quantities --- hyperparameters, observation inputs, location and
number of change points.  This option may, in principle, result
in an efficient and robust implementation, albeit at a high
implementation cost. Another option is to implement the
algorithms in a probabilistic programming framework~\cite{},
which gains support for probabilistic inference on uncertain
parameters but apparently limits implementation efficiency of
Gaussian process-specific computations.

In this work, we advocate a third approach, which combines
advantages of the other two above. We present an architecture of
a Gaussian process library that provides an efficient
implementation of Gaussian process-related computations shared
by all algorithms, but lets program, virtually without
restrictions, both Gaussian process kernels and probabilistic
models of uncertainty, and combine Gaussian processes and
probabilistic models for inference using algorithms supported by
the framework. Our implementation of probabilistic
programming-enabled Gaussian processes, GoGP
(\url{http://bitbucket.org/dtolpin/gogp}), is written in Go
and makes extensive use of modelling and inference capabilities
provided by Infergo~\cite{}, a Go library for differentiable
probabilistic programming. In what follows we first outline GoGP
design  and then illustrate the use of GoGP on a number of case
studies inspired by applications.

@@

Common computationally-intensive algorithms are implemented in 
methods of \lstinline{GP} and rely on efficient parallelism of
the Go language and a highly optimized library for mathematical 
computations Gonum~\cite{}. Problem-specific code comprises two
parts --- a kernel and optionally a probabilistic model --- that
are both automatically differentiated by Infergo.  As a minimum,
a \lstinline{GP} instance must be initialized with a kernel, an
object of type implementing an Infergo model and computing the
covariance between two locations. A few basic kernels are
provided in GoGP as a convenience, and can be combined in derived
kernels using virtually unrestricted Go code.

@@

When observation inputs are uncertain, beliefs about inputs can
be specified, and the log-likelihood gradient can be computed
with respect to both hyperparameters and observation inputs. For
example, in a time series, one can assume that observation
inputs come from a renewal process and let the inputs move 
relative to each other. Then, forecasting can be performed 
relative to posterior observation inputs.

@@

In basic usage, the observation noise is assumed to be Gaussian.
This usage is supported by initializing \lstinline{GP} with a noise
kernel, along with a similarity kernel. When the noise is not
Gaussian, an analytical solution for posterior Gaussian process
inference does not always exist. However, non-Gaussian noise is
straightforwardly supported by GoGP through supplying a
model for beliefs on observation outputs:
\begin{lstlisting}
type Noise struct {
	Y []float64 // noisy outputs
}
func (m *Noise) Observe(x []float64) float64 {
	// Laplacian noise
	for i := range m.Y {
		ll += Expon.Logp(1/math.Exp(x[0]),
			math.Abs(m.Y[i]-x[1+i]))
	}
	return ll
}
\end{lstlisting}

>>>>

Gaussian processes~\cite{RW05} possess properties that make
them the approach of choice in many applications:
\begin{itemize}
    \item A Gaussian process works with as little or as much
    	data as available.
    \item Non-uniformly sampled observations, missing
    	observations, and observation noise are handled
    	organically.
    \item Libraries for inference with Gaussian processes
    	are readily available, both standalone and within
    	machine learning frameworks, such that Gaussian
    	processes can be included into data processing
    	pipelines and combined with other machine learning
    	algorithms.
\end{itemize}

Gaussian processes are successfully deployed in many settings,
in particular when 
\begin{itemize}
    \item a stationary kernel is sufficient;
    \item kernel hyperparameters can be fixed \textit{a priori} or
    	reliably estimated from observations;
    \item observation noise is constant.
\end{itemize}
Most basic cases are indeed covered by the above assumptions,
and library implementations~\cite{MWN+17,GPB+18,PVG+11} of
Gaussian processes prioritize scalability with number of
observations, rich libraries of combinable kernel functions, and
integration of Gaussian processes with machine learning
pipelines.

There are cases though where noise varies between
observations, structured informative priors must be imposed on
kernel hyperparameters, observation inputs are uncertain, or
observations are best explained by presence of change
points~\cite{ROE+13,SGR04,GOR09,MR11,STR10}.

While many of the above cases are addressed in research
literature, their diversity is beyond feasible support by a
library, and the practitioner has to implement advanced Gaussian
process algorithms on case-by-case basis. One, laborious, option
is to implement the algorithms in a general-purpose programming
language, along with probabilistic inference for uncertain
quantities --- hyperparameters, observation inputs, location and
number of change points.  This option may, in principle, result
in an efficient and robust implementation, albeit at a high
implementation cost. Another option is to implement the
algorithms in a probabilistic programming
framework~\cite{TMY+16,Stan17,GXG18},
which gains support for probabilistic inference on uncertain
parameters but apparently limits implementation efficiency of
Gaussian process-specific computations.

In this work, we advocate a third approach, which combines
advantages of the other two above. We present an architecture of
a Gaussian process library that provides an efficient
implementation of Gaussian process-related computations shared
by all algorithms, but lets program, virtually without
restrictions, both Gaussian process kernels and probabilistic
models of uncertainty, and combine Gaussian processes and
probabilistic models for inference using algorithms supported by
the framework. Our implementation of probabilistic
programming-enabled Gaussian processes, GoGP
(\url{http://bitbucket.org/dtolpin/gogp}), is written in Go
and makes extensive use of modelling and inference capabilities
provided by Infergo~\cite{T19}, a Go library for differentiable
probabilistic programming. In what follows we first outline GoGP
design  and then illustrate the use of GoGP on a number of case
studies inspired by applications.

@@

Common computationally-intensive algorithms are implemented in 
methods of \lstinline{GP} and rely on efficient parallelism of
the Go language and a highly optimized library for mathematical 
computations Gonum~\cite{Gonum}. Problem-specific code comprises two
parts --- a kernel and optionally a probabilistic model --- that
are both automatically differentiated by Infergo.  As a minimum,
a \lstinline{GP} instance must be initialized with a kernel, an
object of type implementing an Infergo model and computing the
covariance between two locations. A few basic kernels are
provided in GoGP as a convenience, and can be combined in derived
kernels using virtually unrestricted Go code.

@@

When observation inputs are uncertain~\cite{MR11}, beliefs about inputs can
be specified, and the log-likelihood gradient can be computed
with respect to both hyperparameters and observation inputs. For
example, in a time series, one can assume that observation
inputs come from a renewal process and let the inputs move 
relative to each other. Then, forecasting can be performed 
relative to posterior observation inputs.

@@

In basic usage, the observation noise is assumed to be Gaussian.
This usage is supported by initializing \lstinline{GP} with a noise
kernel, along with a similarity kernel. When the noise is not
Gaussian~\cite{MWN+17}, an analytical solution for posterior Gaussian process
inference does not always exist. However, non-Gaussian noise is
straightforwardly supported by GoGP through supplying a
model for beliefs on observation outputs:
\begin{lstlisting}
type Noise struct {
	Y []float64 // noisy outputs
}
func (m *Noise) Observe(x []float64) float64 {
	// Laplacian noise
	for i := range m.Y {
		ll += Expon.Logp(1/math.Exp(x[0]),
			math.Abs(m.Y[i]-x[1+i]))
	}
	return ll
}
\end{lstlisting}
